Estimation - Confidence intervals
=================================

christophe@pallier.org

Here, we just show how to get confidence intervals for the proportion or the mean obtained from a sample. For confidence intervalle of comparisons, see the other files.

## Estimating a proportion

```{r simulbinom}
n <- 100
a <- rbinom(n, size=1, prob=1/3)
```

```{r tabulate}
table(a)
prop.table(table(a))
mean(a)
```

### Inferential stats

```{r propci}
prop.test(table(a))
prop.test(table(a))$conf.int
```


## Estimating a mean

```{r simul}
n <- 100
a <- rnorm(n, mean=100, sd=15)
```


## Exploratory graphics

```{r}
par(las=1)
stripchart(a, method='jitter', vertical=TRUE)
abline(h=mean(a), lty=2)
boxplot(a)
hist(a)
rug(a)
plot(density(a))
abline(v=mean(a), lty=2)
rug(a)
```

If the sample is small, you can use a `dotchart`

```{r}
dotchart(a[1:20])
```


## Descriptive stats


```{r descriptive stats}
summary(a)
mean(a)
mean(a[abs(a-mean(a)) < 2*sd(a)]) # after deleting point beyond 2 stddev 
```

## Confidence intervals

```{r}
t.test(a)
t.test(a)$conf.int

require(boot)
sampmean <- function(x, d) { mean(x[d]) }
boota <- boot(a, sampmean, 1000)
boot.ci(boota)
```


