---
title: "intro to programming 5"
author: "Henri Vandendriessche"
date: "12/10/2021"
fontsize: 8pt
output:
  slidy_presentation: default
  beamer_presentation: default

---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## R Markdown
Read a file

print content of a file

number of lines
number of words
store number of words in dictionary

Plot it

Special character ?

Create a file

Read a csv file


"
    Querying and navigating the file system https://automatetheboringstuff.com/2e/chapter10/
    Reading and writing text files.
    Manipulating strings (split, join, ...). (chapter 10 of Swinnen's Apprendre à programmer avec Python 3)

Exercices

    Write a script that prints the first 10 lines of a file
    Write a script that prints the last 10 lines of a file (or the whole file is it is less than 10 lines long).
    Write a script that opens and read a text file, and print all the lines that contain a given target word
    compute the number of the number of words (removing punctuation) in a text file
    compute the number of occurences of each word in a text file
    read a matrix in a text file and return the mean of each row (not using pandas, ...)
    find and list all *.csv files in a folder and its subfolders.
    rename all files in a folder by adding the last modification date and the end of the name (before the extension)
"


reading and writing files
file:///home/henri/Documents/Cours/Info-2/EXERCICES-2/python-exercices-02-doc.html#reading-and-writing-files

https://inforef.be/swi/download/apprendre_python3_5.pdf
