---
title: "intro-to-programming-6"
author: "Henri Vandendriessche"
date: "02/11/2021"
fontsize: 8pt
output:
  beamer_presentation: default
  slidy_presentation: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

So far
========================================================

  - Python, its life, its choice

  - Data types: (integer / float / string / boolean)
  
  - **If**, **For** and **While** loops: 

  - Data collections: (list, tuple, set, dictionary)
    
  - Functions
  
  - Higher order function (Recursive function)
  
  - Read and write files


Today
========================================================

  
  - Object-Oriented Programming in Python
  
  - Classes
  
  - Objects

  - Manage exceptions
  
  - Clean Code


Object-Oriented Programming
========================================================


  - Programming paradigm based of the concept of "objects"
  
  - Objects contains data and code
  
  - 
  
Classes 1/
========================================================
  
  - Classes are templates for objects
  
  - Example: You have a power point templates for all your presentations of articles with 
      + a graphic chart
      + your font 
      + your background image
      + all your slides with generic titles (Introduction, study 1, results study 1...)
      
  - That templates in python is a class. A kind of general recipe that specify an object (in our case a power point presentation for a class)
  
  
  

Classes 2/
========================================================
  
  - Classes are templates for objects
  
  - power point templates
  
  
  
Objects
========================================================

  - Objects are instances on classes
  
  - In our metaphor one object is for example the power point of your class of "introduction to Cognitive psychology"
  
  - That power point has the same features and characteristics that the power point of your class of "introduction to Linguistics" yet the content differs.
  
  
  
Create your class
========================================================



Constructors
========================================================
  

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
